Potential competition in alpine bumble bees is mitigated by imperfect trait-matching and subpartitioning within floral morphotypes

### 2021-03-25 ms_2/ms_2_v2.Rmd
fixed nomenclature again

The role of competition in regulating community assembly and species coexistence has long been a central question in ecology. It has taken on new urgency, however, on account of recent climate change, which threatens to disrupt existing patterns of species interactions [@Tylianakis2008-fi], including competition [@Ettinger2017-pk], with consequences for community composition and function. These effects are likely to be most pronounced along the steep climate gradients of mountain slopes, where rapid climate-induced range shifts are expected to introduce novel competitive interactions within trophic levels and cause spatial or phenological mismatches between organisms and their food resources.

Bumble bees (Hymenoptera: *Bombus* spp.) have become classic model organisms for the study of competition and species coexistence due to their broad and overlapping distributions and their gross similarity in morphology, behavior, and dietary niche [@Plowright1984-lc; @Goulson2008-zy]. In this role, however, they have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble communities [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communities have been shown to be over-dispersed in terms of tongue length or phylogeny, suggesting that competition functions as a community assembly process [@Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx; @Goulson2008-zy]. To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous mosaic of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread “core species”, and the effects of competition are more pronounced in the composition of regional species pools than in that of local assemblages [@Hanski1982-ie]. In some cases, competition for food may simply be precluded by a saturating abundance of floral resources [@Williams1989-vr], implying that populations are limited by other factors, such as abiotic stressors, nest site availability, or predator/pathogen pressure.

Teasing apart the still-obscure mechanisms of bumble bee coexistence and niche partitioning remains a salient problem from a basic ecological perspective, but it has also become an urgent question in the context of applied conservation. Bumble bees are cold-adapted species [@Heinrich1994-cv] associated mainly with mountain habitats, where they serve as the principal pollinators of many alpine plants [@Williams1998-xi; @Goulson2010-lo]. Warming climates have already caused measurable upslope shifts in bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], and these are predicted to cause declines in many bumble bees species [@Soroye2020-zb]. Moreover, the introduction or extinction of species from bumble bee communities, with corresponding addition and release of competitive pressure, can alter both the floral visitation patterns of bumble bees and, consequently, the pollination biology of their host plants [@Brosi2013-mu; @Ishii2013-sm]. Thus, the speculative question of how competition has acted in the historic assembly of bumble bee communities has become an unfolding process observable over ecological time scales and instrumental in the future viability of bumble bee species and their floral mutualists.

All else held equal, species coexistence is favored when intraspecific competition exceeds interspecific competition, and this stabilizing inequality is favored to the extent that species differ in dietary niche [@Abrams1977-zp; @Chesson2000-aw]. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs associated with exploiting it. These costs are closely related to the degree of morphological fittedness between bumble bees and their floral hosts, which consists principally in the matching of bumble bee tongue length to floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], which leads to an association between tongue length and floral selection in bumble bee communities [e.g. @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. This association between tongue length and floral niche, though, is imperfect, and there can be considerable overlap in floral resource use between bumble bees of different tongue lengths [e.g. @Ranta1980-ok; @Goulson2008-oe].

Empirical measurements of niche overlap are, however, inherently ambiguous with respect to competition. It has been well-demonstrated that niche overlap cannot serve as a metric of competition *per se* [@Abrams1980-am], but it is often used as an index of *potential* competition, particularly in the context of dietary studies [e.g. @French1985-eo; @Steffan-Dewenter2000-ll; @Jung2015-cw], based on the reasoning that if and when food becomes limiting, competition can be expected to vary in proportion to the dietary overlap of co-occurring species [@Pianka1972-oi]. Focusing on *potential* competition does not, however, escape the equivocity of niche overlap, since an observed pattern of niche overlap could just as plausibly be interpreted as adaptive niche partitioning in response to *realized* competition [@Pianka1974-qk]. This creates a predicament in which niche overlap can only be interpreted as an index of potential competition when realized competition is assumed to be negligible, an assumption that can only be justified on the basis of additional knowledge of the study system under consideration.

Our study is set in the Berchtesgadener Alps of southeast Germany, a landscape of flower-rich mountain meadows embedded in a matrix of coniferous forest. This system is one of rich floral resources but harsh abiotic conditions, so it is likely that competition for floral resources is low, or at least transient, and that bumble bees are limited instead by the shortness of the growing season and by fluctuating abiotic conditions during the critical period of spring colony founding. On this basis, we argue for the interpretation of niche overlap as an index of *potential* competition rather than a response to *realized* competition, with the caveat this interpretation may misrepresent certain temporal intervals (e.g. early spring) or elevational bands (e.g. the sparse alpine grasslands above the tree line ecotone) [Sponsler et al., submitted].

The aim of our study is to synthesize the phenomena of floral preference, trait-matching, and dietary niche overlap in a way that accounts for the coexistence of bumble bee species with similar tongue lengths and sheds light on the possible effects of climate-driven elevational range shifts on the future of potential competition between bumble bee species in mountain habitats. We first ask how floral preferences compare across bumble bee species and to what extent these patterns can be explained by trait-matching between bumble bee tongue length and floral morphotype. We then investigate the patterns of dietary niche overlap between bumble bee species that emerge from their respective floral preferences, comparing niche overlap at the nested levels of floral morphotype, floral family, and floral genus. Finally, we ask how niche overlap varies with elevation and discuss the implications of this variation for bumble bee conservation in a changing climate.

@@

<!-- I guess I'll need to tweak the wording of this section so that it is not a verbatim copy of the other manuscript. But it's okay if it recycles a lot of language because it is, after all, the same thing being described. -->

<!-- Describe general floristic and temperature patterns. Describe sampling in more detail. -->

@@

Sampling consisted of repeated visits (hereafter, “visitation samples”) to each study site at approximately weekly intervals. Samples were only collected on dry days when the air temperature was at least 6°C. During each visit, bumble bees and their activity were recorded during a 50-minute transect walk. Bumblebees observed on a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, whereas workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory after @Amiet1996-jz. While floral visitation by male bumble bees was recorded during visitation sampling, we chose to analyze only visitation by queens and workers, which we pooled. In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.1 m^2^. Plant species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

While per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) varied markedly, with 153 in 2010, 212 in 2011, and 539 in 2012. The increased total sampling effort in 2012 was driven in part by the overall scarcity of bumble bees that year. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient.

@@

## Software

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. Bayesian multilevel models were implemented in Stan [@Stan2020aa] via the package `brms` [@Burkner2017-aa] using default priors, and visualizations were prepared with the packages `tidybayes` [@Kay2020-aa] and `ggdist` [@Kay2021-aa]. Ordination analysis and permutation testing were performed with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Hierarchical generalized additive models (HGAMs) were implemented with the package `mgcv` [@Wood_2017] and visualized with the packages `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta-aa2021]. All `R` code used in this analysis, together with full model output and diagnostics, is available in Appendix X.

@@

We calculated niche overlap with the goal of focusing on elevational patterns, sacrificing temporal resolution to pool observations within site. First, we calculated each bumble bee species' proportional visitation frequency at the nested levels of floral morphotype, plant family, and plant genus. Then, we calculated, for each bumble bee species, the mean proportional visitation rate to each morphotype/family/genus across all sampling dates within each site. Using these species-level visitation matrices, we calculated pairwise niche overlap as 1 - Horn-Morisita distance (calculated separately for morphotype-, family-, and genus-level matrices) for all species pairs within each site. This approach was chosen for its normalizing effect on the highly variable "snapshot" visitation data generated by our individual transect walks, but it entails the disadvantage of conflating niche overlap due to similar floral preference and niche overlap due to similar temporal (i.e. seasonal) activity patterns. We opted to analyze niche overlap only down to the level of floral genus in order to dampen the possible effects of false non-detection of floral species due to undersampling of the diets of rare bumble bee species [@Williams2005-yp].

While our study spanned three years, we chose not to model year effects. Differences in sampling intensity and duration across years (described above) would make comparisons across years difficult to interpret. This means that our analysis cannot detect year-to-year variation in floral biases and niche overlap, and main effects cannot be distinguished from annual anomalies. With only three years of data, though, such annual inferences would be of little value, and we deemed them not worth the cost of added model complexity.

@@

@@

We first compared overall visitation probability across floral morphotypes. In this model, visitation for each sample was pooled across bumble bee species and within floral morphotypes prior to binarization. Visitation probability was then modeled using multilevel Bayesian regression with a Bernoulli distribution family (logit link function). *Bumble bee abundance* and *percent flower cover* (per morphotype) were included as a global interaction term, thus conditioning visitation probability on abundance and enabling the inference of preference independent of the neutral effects of abundance. *Floral morphotype* was specified as a both a varying intercept term and a varying slope term with respect to *percent flower cover*. *Site* was included as a varying intercept term to avoid pseudoreplication owing to the repeated observations within each site.

@@

To explore the functional underpinnings of bumble foraging preference, we analyzed visitation probability as a function of the interaction between floral morphology and bumble bee tongue length class. For this analysis, visitation was pooled within floral morphotype but not across bumble bee species. We then specified a Bayesian model of the same form as Model 1, but this time included a varying slope and intercept term for floral morphotype : bumble bee tongue length class and floral morphotype : bumble bee species. In this model, the relationship between floral morphotype and bumble bee tongue length class was the focus of inference, and the bumble bee species effect was added to account for the grouping of observations within species.

@@

To visualize the overall diet similarity between bumble bee species, we pooled visitation data across all samples within bumble bee species and plant genus, then performed a principal coordinates analysis based on Horn-Morisita distance metric. To test the significance of tongue length class as a predictor of diet similarity, we performed a permutation test. Importantly, this approach treats the whole study system in aggregate, thus conflating the causal effects of floral preference, seasonal activity patterns, and site occupancy patterns in the generation of overall diet similarity. In this sense, it can be understood as a model of potential diet overlap given the aggregate foraging patterns of the bumble bee species in our system, with the understanding that realized niche overlap at particular locations or times may vary.

@@

To investigate elevational patterns of *within-site* niche overlap, we modeled community-mean niche overlap as a function of level (morphotype, family, genus) and elevation using HGAMs. Community mean niche overlap, which we interpret as an index of potential *diffuse* competition [@Pianka1974-qk; @MacArthur1972-qz], was calculated by taking the mean niche overlap (within level) across all species pairs for each site. Level was included as both a fixed intercept effect and a smoothing factor in conjunction with elevation, following the type "I" model form described by @Pedersen2019-kf. To account for the proportional nature of our data, we used a beta distribution family with a logit link function.

@@

To focus on genus-level niche overlap as a function of tongue length class, we generated a binary tongue-difference variable for each species pair: "same" for species belonging to the same tongue length class and "different" for species belonging to different tongue length classes. We then took the mean genus-level niche overlap across all species pairs within tongue-difference class and site. Then, we modeled genus-level niche overlap using tongue-difference as both a fixed intercept effect and a smoothing factor in conjunction with elevation, again following the type "I" model form @Pedersen2019-kf.

@@

We recorded a total of 12,918 bumble- bee-flower interactions over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species (with *Bombus terrestric/lucorum* and *Bombus psithyrus* species lumped into species groups), 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily above 1500 m, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across mid-elevation sites. *B. hypnorum* and *B. humilis* were recorded only three times and only in one year, so we omitted it from all analyses.

Floral surveying yielded a total of 352 plant species, representing 191 genera and 52 families. Of these, 155 species from 103 genera and 35 families were observed to be visited by bumble bees. Eight species --- *Rubus idaeus*, *Rosa canina*, *Juniperus communis*, *Larix decidua*, *Salix* sp., *Caltha palustris*, *Pulmonaria officinalis*, and *Rheum barbarum* --- were recorded in visitation data but not recorded during floral surveying. Each accounted for no more than 4 visits in total over the three years of our study, and they were omitted from the analysis of floral abundance.

@@

Lip (group 5) and flag (group 6) flowers had the highest overall visitation probability, with a conditional mean visitation probabilities of 0.35 (95% CI = 0.27–0.4) and 0.31 (95% CI = 0.24–0.40), respectively (Figure 1). In second place, head flowers (group 7) had a mean conditional visitation probability of 0.18 (95% CI = 0.13–0.25). Bell (group 3) flowers were a distant third with a conditional mean visitation probability of 0.09 (95% CI = 0.06–0.13), followed by stalk-disc (group 4), funnel (group 2), and disc flowers (group 1) with conditional visitation probabilities of 0.06 (95% CI = 0.04–0.08), 0.04 (95% CI = 0.03–0.07), and 0.04 (95% CI = 0.02–0.06), respectively. Model 1 explained approximately 37% of the total variance (97.5% CI = 35–38%).

```{r fig1, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by floral morphotype. Point-interval plots depict 66\\% and 95\\% CIs. Each floral morphotype is exemplified by a representative species common in our study system. Illustrations are modifications of public domain works obtained from www.plantillustrations.org. For details of original sources, see Appendix X.")}
knitr::include_graphics("../output/brm_pvis01_plot1_mod.png")
```

Disc (group 1) flowers were visited at similar rates across tongue length classes **(Figure 2)**. At the species level, disc flowers were visited at relatively high rates by the short-tongued *B. terrestris/lucorum* and the medium-tongued *Bombus monticola* **(Figure 3)**. Funnel (group 2) flowers were preferred by long-tongued species, an effect driven by the *Aconitum* (group 2) specialist *B. gerstaeckeri*. Bell (group 3) flowers were visited at a higher rate by short-tongued species than by long-tongued species, with medium-tongued species exhibiting intermediate visitation rates. At the species-level, bell flowers were most preferred by the short-tongued species *B. jonellus*, *B. pyrenaeus*, and *B. soroeensis*, and by the medium-tongued species *B. hypnorum* and *B. monticola*. Stalk-disc (group 4) flowers were visited at similar rates across all tongue length classes, and there was also broad overlap in visitation rates at the species level. Lip (group 5) flowers were also visited at similar rates across tongue length classes, but at the species level *B. pascuorum* exhibited an uniquely high preference for lip flowers. Flag (group 6) flowers were visited at a higher rate by medium-tongued species than by short-tongued species, with long-tongued species exhibiting intermediate visitation rates. Flag flowers generated strong separation in visitation rate at the species level, receiving very high visitation from the medium-tongued *B. pascuorum* and *B. mucidus*, moderately high visitation from the short-tongued *B. terrestris/lucorum* and *B. lapidarius*, the medium-tongued *B. monticola*, and the long-tongued *B. mendax* and *B. hortorum*, and low visitation by the short-tongued *B. soroeensis*, *B. pyrenaeus*, *B. psithyrus*, *B. pratorum*, and *B. jonellus*, the medium-tongued *B. hypnorum*, and the long-tongued *B. gerstaeckeri*. Models 2 and 3 both explained approximately 31% of total variance (97.5% CI = 30–32%).

```{r fig3, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability parsed by Kugler morphotype and bumble bee tongue length class (Model 2). Point-interval plots depict 66\\% and 95\\% CIs. Letters indicate group differences based on evidence ratio at alpha = 0.05. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across tongue length classes within floral morphotypes.")}
knitr::include_graphics("../output/brm_pvis00_plot1.png")
```

>>>>

The role of competition and niche partitioning in regulating community assembly has long been a central question in ecology, but it has taken on new urgency on account of anthropogenic climate and land use change, which threaten to disrupt existing patterns of species interactions and niche relationships [@Tylianakis2008-fi]. These effects are likely to be especially pronounced in mountain ecosystems, where steep elevational temperature gradients facilitate species range shifts in response to climate change, and socioeconomic factors are driving rapid changes in historical land use patterns [@Becker2007-cg; @Loffler2011-xo; @Klein2019-dq]

Bumble bees (Hymenoptera: *Bombus* spp.) have become classic model organisms for the study of competition and species coexistence due to their broad and overlapping distributions and their gross similarity in morphology, behavior, and dietary niche [@Plowright1984-lc; @Goulson2008-zy]. In this role, however, they have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble communities [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communities have been shown to be overdispersed in terms of tongue length or phylogeny, suggesting that competition functions as a community assembly process [@Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx; @Goulson2008-zy]. To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous mosaic of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread “core species”, and the effects of competition are more pronounced in the composition of regional species pools than in that of local assemblages [@Hanski1982-ie]. In some cases, competition for food may simply be precluded by a saturating abundance of floral resources, implying that populations are limited by other factors, such as abiotic stressors, nest site availability, or predator/pathogen pressure [@Ranta1981-vx; @Williams1989-vr].

All else held equal, species coexistence is favored when intraspecific competition exceeds interspecific competition, and this stabilizing inequality is favored to the extent that species differ in dietary niche [@Abrams1977-zp; @Chesson2000-aw]. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs associated with exploiting it. These costs are closely related to the degree of morphological fittedness between bumble bees and their floral hosts, which consists principally in the matching of bumble bee tongue length to floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], which leads to an association between tongue length and floral choice in bumble bee communities [e.g. @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. This association between tongue length and floral niche is imperfect, though, and there can be considerable overlap in floral resource use between bumble bees of different tongue lengths [e.g. @Ranta1980-ok; @Goulson2008-oe].

While niche overlap cannot serve as a metric of competition *per se* [@Abrams1980-am], it is often used as an index of *potential* competition, particularly in the context of dietary studies [e.g. @French1985-eo; @Steffan-Dewenter2000-ll; @Jung2015-cw], based on the reasoning that if and when food becomes limiting, competition can be expected to vary in proportion to the dietary overlap of co-occurring species [@Pianka1972-oi]. Our study is set in the Berchtesgadener Alps of southeast Germany, a landscape of flower-rich mountain meadows embedded in a matrix of coniferous forest. This system is one of abundant floral resources but harsh abiotic conditions, so it is likely that competition for floral resources is low, or at least transient, and that bumble bee populations are limited instead by the shortness of the growing season. Yet while *realized* competition between bumble bees in our system may be low, the question of *potential* competition is highly relevant because mountain ecosystems are currently subject to at least two powerful drivers of directed change. First, warming climates have already caused measurable upslope shifts in bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], which could introduce novel competitive interactions between bumble bee species that experience increased sympatry. Second, socioeconomic factors have led to a dramatic decline in traditional alpine grazing management, particularly in the European Alps, resulting in widespread pasture abandonment and reforestation of mountain meadows [@Tappeiner2006-jp; @Tasser2007-qo]. The regression of meadows to forest entails a massive loss of floral resource availability, and even early stages of succession can cause declines in floral abundance and diversity, with corresponding effects on bumble bees [@Walcher2017-kw]. At the same time, more accessible areas at the bases of mountains are undergoing agricultural intensification that displaces natural vegetation and favors grasses over forbs [@Tasser2002-om]. Caught between adverse land use changes above and below and subject to climate-induced range shifts, mountain bumble bees are facing a future of reduced floral resource availability and altered sympatry, and potential competition is likely to become realized according to patterns of niche overlap between species.

Thus, the still-obscure mechanisms of bumble bee coexistence and niche partitioning remain a salient problem from a basic ecological perspective, but they have also become timely questions in the context of applied conservation. The aim of our study is to synthesize the phenomena of floral preference, trait-matching, and dietary niche overlap in a way that accounts for the coexistence of bumble bee species with similar tongue lengths and sheds light on the possible effects of habitat loss and climate-driven range shifts on the future of competition between bumble bee species in mountain habitats. We first ask how floral preferences compare across bumble bee species and to what extent these patterns can be explained by trait-matching between bumble bee tongue length and floral morphotype. We then investigate the patterns of dietary niche overlap between bumble bee species that emerge from their respective floral preferences, comparing niche overlap at the nested levels of floral morphotype, floral family, and floral genus. Finally, we ask how niche overlap varies with elevation and discuss the implications of this variation for bumble bee conservation in an era of global change.

@@

@@

Sampling consisted of repeated visits (hereafter, “visitation samples”) to each study site at approximately weekly intervals. Sampling was only conducted during periods without rain and with an air temperature of at least 6°C. Bumble bees were recorded during one 50-minute transect walk per visit, and bumblebees observed on a given flower were counted as floral visitors. Queens were identified to species level in the field to avoid destructively sampling colony foundresses, but workers and males were collected and frozen for later identification in the laboratory. Species were identified following @Amiet1996-jz. While male bumble bees was recorded, we chose to analyze only the pooled visitation of by queens and workers. In conjunction with bumble bee observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.01 m^2^ and identified the plant species according to @Lauber2007-ha and @Oberndorfer2001-ro.

While per-sample effort was standardized across all years, the total number of samples varied: 153 in 2010, 212 in 2011, and 539 in 2012. The start of field sampling each year was determined by snowmelt, so lower sites were sampled as early as April while higher sites could not be sampled until June or July. In 2010 and 2012, this resulted in a temporally staggered pattern of sampling, but a combination of early snowmelt and a late start to field activities in 2011 resulted in sampling that was approximately aligned in time across the whole elevation gradient. Due to these annual differences in sampling intensity and duration, we opted to pool data across years rather than include year effects in our models, which would multiply model complexity and be difficult to interpret. This means, however, that our analysis cannot detect year-to-year variation in floral preferences and niche overlap, and main effects cannot be distinguished from annual anomalies. With only three years of data, though, such annual inferences would be of little value, and we deemed them not worth the cost of added model complexity.

The samples analyzed in this paper have been previously analyzed (Sponsler et al., submitted), but the analyses presented here are non-overlapping.

@@

@@

We calculated niche overlap with the goal of focusing on elevational patterns, sacrificing temporal resolution to pool observations within site. First, we calculated each bumble bee species' proportional visitation frequency at the nested levels of floral morphotype, plant family, and plant genus. Then, we calculated, for each bumble bee species, the mean proportional visitation rate to each morphotype/family/genus across all sampling dates within each site. Using these species-level visitation matrices, we calculated pairwise niche overlap as 1 - Horn-Morisita distance (calculated separately for morphotype-, family-, and genus-level matrices) for all species pairs within each site. This approach was chosen for its regularizing effect on the highly variable "snapshot" visitation data generated by our individual transect walks, but it entails the disadvantage of conflating niche overlap due to similar floral preference and niche overlap due to similar temporal (i.e. seasonal) activity patterns. We opted to analyze niche overlap only down to the level of floral genus in order to dampen the possible effects of false non-detection of floral species due to undersampling of the diets of rare bumble bee species [@Williams2005-yp].

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. All `R` code used in this analysis, together with full model output and diagnostics, is available in Appendix X.

@@

We first compared overall visitation probability across floral morphotypes. In this model, visitation for each sample was pooled across bumble bee species and within floral morphotypes prior to binarization. Visitation probability was then modeled using multilevel Bayesian regression with a Bernoulli distribution family (logit link function). Models were implemented in Stan [@Stan2020aa] via the package `brms` [@Burkner2017-aa], and visualizations were prepared with the packages `tidybayes` [@Kay2020-aa] and `ggdist` [@Kay2021-aa]. *Bumble bee abundance* and *percent flower cover* (per morphotype) were included as a global interaction term, thus conditioning visitation probability on abundance and enabling the inference of preference independent of the neutral effects of abundance. *Floral morphotype* was specified as a both a varying intercept term and a varying slope term with respect to *percent flower cover*. *Site* was included as a varying intercept term to avoid overfitting due to pseudoreplication owing to the repeated observations within each site. We set a weakly informative (i.e. regularizing) Normal(0, 1.5) prior on the β-coefficients of the fixed effects [@Lemoine2019-xm] and used the default priors provided by `brms` version 2.14.4 for all other model parameters.

@@

To explore the functional underpinnings of bumble foraging preference, we analyzed visitation probability as a function of the interaction between floral morphology and bumble bee tongue length class. For this analysis, we pooled visitation within floral morphotype but not across bumble bee species. Then we specified a Bayesian model of the same form as Model 1, but this time included a varying slope and intercept term for *floral morphotype* : *bumble bee tongue length class* and *floral morphotype* : *bumble bee species*. In this model, the relationship between *floral morphotype* and *bumble bee tongue length class* was the focus of inference, and the *bumble bee species* effect was added to account for the grouping of observations within species.

@@

To visualize the overall diet similarity between bumble bee species, we pooled visitation data across all samples within bumble bee species and plant genus, then performed a principal coordinates analysis (PCoA) based on Horn-Morisita distance metric. To test the significance of tongue length class as a predictor of diet similarity, we performed a permutation test. Importantly, this approach treats the whole study system in aggregate, thus conflating the causal effects of floral preference, seasonal activity patterns, and site occupancy patterns in the generation of overall diet similarity. In this sense, it can be understood as a model of potential diet overlap given the aggregate foraging patterns of the bumble bee species in our system, with the understanding that realized niche overlap at particular locations or times may vary. PCoA and permutation tests were performed with the R package `vegan` [@Oksanen_2019], and visualizations were prepared with the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]

@@

To investigate elevational patterns of *within-site* niche overlap, we modeled community-mean niche overlap as a function of level (morphotype, family, genus) and elevation using hierarchical generalized additive models (HGAMs). Community mean niche overlap, which can be interpreted as an index of potential *diffuse* competition [@Pianka1974-qk; @MacArthur1972-qz], was calculated by taking the mean niche overlap (within level) across all species pairs for each site. Level was included as both a fixed intercept effect and a smoothing factor in conjunction with elevation, following the type "I" model form described by @Pedersen2019-kf. Management was initially included as a fixed intercept effect, but its influence was insiginificant, so it was dropped from the final model. To account for the proportional nature of our data, we used a beta distribution family with a logit link function. Models were fit with the R package `mgcv` [@Wood_2017] and visualized with the packages `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta-aa2021].

@@

To focus on genus-level niche overlap as a function of tongue length class, we generated a binary tongue-difference variable for each species pair: "same" for species belonging to the same tongue length class and "different" for species belonging to different tongue length classes. We then took the mean genus-level niche overlap across all species pairs within tongue-difference class and site. Then, we constructed an HGAM of genus-level niche overlap using tongue-difference as both a fixed intercept effect and a smoothing factor in conjunction with elevation, again following the type "I" model form @Pedersen2019-kf.

@@

We recorded a total of 12,918 bumble- bee-flower interactions, including 16 bumble bee species (with *Bombus terrestric/lucorum* and *Bombus psithyrus* species lumped into species groups), 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii* were the most abundant species, and they occurred at all sites. *B. hortorum*, *B. jonellus*,  *B. monticola*, and *B. psithyrus* occurred at moderate abundance. The remaining species --- *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius*, *B. mendax*, *B. mucidus*, and *B. pyrenaeus* --- were relatively rare. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses. Floral surveying yielded a total of 352 plant species (191 genera, 52 families), of which 155 (103 genera, 35 families) observed to be visited by bumble bees. For a more full analysis of bumble bee and floral abundance patterns, see Sponsler at al. (submitted).

@@

Lip and flag flowers were the most preferred morphotypes overall, with similarly high visitation probabilities of 0.35 (95% CI = 0.27–0.44) and 0.31 (95% CI = 0.24–0.40), respectively **(Figure 1)**. In second place, head flowers had a visitation probability of 0.18 (95% CI = 0.13–0.25). Bell flowers were a distant third with a visitation probability of 0.09 (95% CI = 0.06–0.13), followed by stalk-disc, funnel, and disc flowers with visitation probabilities of 0.06 (95% CI = 0.04–0.08), 0.04 (95% CI = 0.03–0.06), and 0.04 (95% CI = 0.02–0.06), respectively. Model 1 explained approximately 37% of the total variance (97.5% CI = 35–38%).

```{r fig1, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by floral morphotype. Point-interval plots depict 66\\% and 95\\% CIs. Each floral morphotype is exemplified by a representative species common in our study system. Illustrations are modifications of public domain works obtained from www.plantillustrations.org. For details of original sources, see Appendix X.")}
knitr::include_graphics("../output/brm_pvis_01_plot1_mod.png")
```

Disc and stalk-disc flowers were visited at similarly low rates across tongue length classes, and lip flowers were visited at similarly high rates across tongue length classes **(Figure 2)**. There was considerable in visitation rates between individual species, though, including those belonging to the same tongue length class **(Figure 3)**. The short-tongued *B. terrestris/lucorum*, for example, stands out in its preference for disc flowers, and the medium-tongued *B. monticola* in its preference for stalk-disc flowers. While lip flowers were visited at relatively high rates by almost all species, the medium-tongued *B. pascuorum* exhibited an especially high preference for this morphotype, while the medium-tongued *B. mucidus* ignored lip flowers almost entirely.

In contrast, visitation of funnel, bell, flag, and head flowers differed significantly both across tongue length classes **(Figure 2)** and between individual species **(Figure 3)**. Funnel flowers were preferred by long-tongued species, though effect was driven mainly by the *Aconitum* specialist *B. gerstaeckeri*. Bell flowers were visited at a higher rate by short-tongued species than by long-tongued species, with medium-tongued species exhibiting intermediate visitation rates. At the species-level, bell flowers were most preferred by the short-tongued species *B. jonellus*, *B. pyrenaeus*, and *B. soroeensis*, and by the medium-tongued species *B. hypnorum* and *B. monticola*. Flag flowers were visited at a higher rate by medium-tongued species than by short-tongued species, with long-tongued species exhibiting intermediate visitation rates. Flag flowers generated strong separation in visitation rate at the species level, receiving very high visitation from the medium-tongued *B. pascuorum* and *B. mucidus*, moderately high visitation from the short-tongued *B. terrestris/lucorum* and *B. lapidarius*, the medium-tongued *B. monticola*, and the long-tongued *B. mendax* and *B. hortorum*, and low visitation by the short-tongued *B. soroeensis*, *B. pyrenaeus*, *B. psithyrus*, *B. pratorum*, and *B. jonellus*, the medium-tongued *B. hypnorum*, and the long-tongued *B. gerstaeckeri*. Models 2 and 3 both explained approximately 31% of total variance (97.5% CI = 30–32%).

```{r fig2, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability parsed by Kugler morphotype and bumble bee tongue length class (Model 2). Point-interval plots depict 66\\% and 95\\% CIs. Letters indicate group differences based on evidence ratio at alpha = 0.05. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across tongue length classes within floral morphotypes.")}
knitr::include_graphics("../output/brm_pvis_00_plot1_mod.png")
```
